Fast gravitational wave parameter estimation without compromises

### 2023-08-01 src/tex/ms.tex
put back em text and edited ROQ text

% Old codes have trouble handling next gen data
% The PE results in the most recent catalog of GW events are produced by several
% community-developed PE codes
% \mi{what catalog do you mean? The LVK doesn't use PyCBC Inference}
There exist a number of prominent, community-developed PE codes, including
\textsc{LALInference} \citep{Veitch:2014wba}, \textsc{PyCBC Inference}
\citep{Biwer:2018osg}, and \textsc{Bilby}
\citep{Ashton:2018jfp,Romero-Shaw:2020owr}.  These packages have been tested by a
number of groups and are well regarded as standard tools. However, while these
tools have passed many robustness tests, they are known to be computationally
intensive. The exact amount of time needed to analyze one event depends on
factors like the duration and frequency of the signal, as well as features of
the specific waveform model. Typical runtimes for production-level analyses can
range from hours to weeks.

@@

To demonstrate the performance of our parameter estimation pipeline, we apply it
to two real LIGO-Virgo events: GW150914 and GW170817. We use the priors shown in
Table \ref{tab:parameters}, and take 4 s of data sampled at 2048 Hz starting at
20 Hz for GW150914, and 128 s of data sampled at 4096 Hz starting at 23 Hz for
GW170817; strain data and PSDs for both events are fetched from GWOSC
\citep{GWOSC}. We use the same normalizing flow model as the injection-recovery
study. For our specific choice of sampler settings, we produce ${\sim}$2500 and
3500 \emph{effective samples}\footnote{Effective samples here refers to the
number of independent samples, which is the total number of generated samples
divided by their correlation length; we compute the effective sample size using
\textsc{arviz} \citep{arviz_2019},
\url{https://python.arviz.org/en/stable/api/generated/arviz.ess.html}. } for
GW150914 and GW170817 respectively. \corr{Running on a node with an
\textsc{Nvidia} 80GB A100 GPU and 16 intel skylake CPUs} \footnote{Note that the
CPU time is hard to benchmark in \textsc{Jax} due to JIT compilation and
asynchronous dispatch. Overall, we find 4 CPUs to be sufficient for our use
case. We chose 16 because of the ease of use of the local cluster.}, the wall
time for both events is around 3 minutes. Most of this time is spent on
just-in-time (JIT) compilation of the code; the actual sampling time is only
${\sim}40$ s. We pre-compute the reference waveform parameters used to
hetetrodyne the likelihood for the two events, which is omitted in the wall-time
calculation. \corr{In terms of specific configuration of the sampler, we used
1000 chains for both of the events. We run both the tuning phase and the
production phase for 20 loops each. Within each phase, the global sampler and
local sampler are stepped 200 times. This translates into $1.6 \times 10^7$
likelihood evaluations in total.\footnote{While in principle adding more
chains makes the algorithm less efficient in terms of the acceptance/evaluation
ratio, since GPUs can evaluate a fixed number of instructions per cycle, we can
keep adding chains without paying extra computational cost before we saturate
the GPU. This results in an improvement in wall time while being less efficient.
We decide to choose a setting that improves the wall time instead of efficiency.} The
acceptance rate of the global proposal is about $2\%$ and $5\%$ for GW150914 and
GW170817 respectively. Note that the setting is chosen to run longer than
necessary to ensure convergence; we are currently pursing tuning studies to
understand trade-offs in the algorithms and improve the efficiency further.}

@@

For comparison, we produce equivalent runs with \textsc{Bilby}, using the same
exact data and priors.  We use the \textsc{dynesty} sampler
\citep{2020MNRAS.493.3132S,dynesty}, with 1000 live points and other settings as
in the configurations files available on github
\url{https://github.com/kazewong/TurboPE/tree/main/src}.  We carry out these
runs using \textsc{parallel Bilby} (\textsc{pBilby}) \citep{Smith:2019ucc} to
distribute the computation over 400 Intel Skylake CPUs for each event. For the
specific settings chosen, the wall-time duration of each run was ${\sim}2$ h for
GW150914 and ${\sim}1$ day for GW170817. \corr{We have also benchmarked the
performance using a ROQ likelihood, which takes about 30 minutes for GW170817.
Without ROQ, it takes about a week to complete a run. This broadly agrees with
what is suggested in the literature \citep{Smith:2016qas}}

>>>>

% Old codes have trouble handling next gen data
% The PE results in the most recent catalog of GW events are produced by several
% community-developed PE codes
% \mi{what catalog do you mean? The LVK doesn't use PyCBC Inference}
There exist a number of prominent, community-developed PE codes, including
\textsc{LALInference} \citep{Veitch:2014wba}, \textsc{PyCBC Inference}
\citep{Biwer:2018osg}, and \textsc{Bilby}
\citep{Ashton:2018jfp,Romero-Shaw:2020owr}.  These packages have been tested by a
number of groups and are well regarded as standard tools. However, while these
tools have passed many robustness tests, they are known to be computationally
intensive. The exact amount of time needed to analyze one event depends on
factors like the duration and frequency of the signal, as well as features of
the specific waveform model. Typical runtimes for production-level analyses can
range from hours to weeks.
This expense precludes iterating quickly on results,
launching large scale measurement simulations, or obtaining final parameter estimates in low
latency to inform astronomers for potential followup in real time.%
\footnote{\corr{Although approximate skymaps can currently be produced in low latency, the same is not true 
of final estimate of other parameters, like masses and spins, which may inform the probability of electromagnetic emission
 followup strategy and thus affect followup strategy.}}

@@

To demonstrate the performance of our parameter estimation pipeline, we apply it
to two real LIGO-Virgo events: GW150914 and GW170817. We use the priors shown in
Table \ref{tab:parameters}, and take 4 s of data sampled at 2048 Hz starting at
20 Hz for GW150914, and 128 s of data sampled at 4096 Hz starting at 23 Hz for
GW170817; strain data and PSDs for both events are fetched from GWOSC
\citep{GWOSC}. We use the same normalizing flow model as the injection-recovery
study. For our specific choice of sampler settings, we produce ${\sim}$2500 and
3500 \emph{effective samples}\footnote{Effective samples here refers to the
number of independent samples, which is the total number of generated samples
divided by their correlation length; we compute the effective sample size using
\textsc{arviz} \citep{arviz_2019},
\url{https://python.arviz.org/en/stable/api/generated/arviz.ess.html}. } for
GW150914 and GW170817 respectively. \corr{Running on a node with an
\textsc{Nvidia} 80GB A100 GPU and 16 intel skylake CPUs} \footnote{Note that the
CPU time is hard to benchmark in \textsc{Jax} due to JIT compilation and
asynchronous dispatch. Overall, we find 4 CPUs to be sufficient for our use
case. We chose 16 because of the ease of use of the local cluster.}, the wall
time for both events is around 3 minutes. Most of this time is spent on
just-in-time (JIT) compilation of the code; the actual sampling time is only
${\sim}40$ s. We pre-compute the reference waveform parameters used to
hetetrodyne the likelihood for the two events, which is omitted in the wall-time
calculation.

\corr{We configured our sampler to use
1000 chains for both of the events. and run both the tuning and production phasesfor 
20 loops each. Within each phase, the global sampler and
local sampler are stepped 200 times, which translates into $1.6 \times 10^7$
likelihood evaluations in total.\footnote{While in principle adding more
chains makes the algorithm less efficient in terms of the acceptance/evaluation
ratio, since GPUs can evaluate a fixed number of instructions per cycle, we can
keep adding chains without paying extra computational cost before we saturate
the GPU. This results in an improvement in wall time while being less efficient.
We decide to choose a setting that improves the wall time instead of efficiency.} The
acceptance rate of the global proposal is about $2\%$ and $5\%$ for GW150914 and
GW170817 respectively. The settings are chosen conservatively to run longer than
necessary to ensure convergence; we are currently pursing tuning studies to
understand trade-offs in the algorithms and improve the efficiency further.}

@@

For comparison, we produce equivalent runs with \textsc{Bilby}, using the same
exact data and priors.  We use the \textsc{dynesty} sampler
\citep{2020MNRAS.493.3132S,dynesty}, with 1000 live points and other settings as
in the configurations files available on github
\url{https://github.com/kazewong/TurboPE/tree/main/src}.  We carry out these
runs using \textsc{parallel Bilby} (\textsc{pBilby}) \citep{Smith:2019ucc} to
distribute the computation over 400 Intel Skylake CPUs for each event. For the
specific settings chosen, the wall-time duration of each run was ${\sim}2$ h for
GW150914 and ${\sim}1$ day for GW170817. \corr{We have also benchmarked
performance using an ROQ likelihood, which takes about 30 minutes for GW170817 withour parallelization;
without ROQ, the same analysis would take about one week to complete. This broadly agrees with
the ROQ performance boost reported in the literature \citep{Smith:2016qas}}
